Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
moving licensing discussion to cross-cutting

This report attempts to summarize the various aspects of the workshop.
The remainder of this paper first gives an overview of the process
with which the workshop was put together (\S\ref{sec:process}), then
proceeds with summaries of the two keynote
presentations~(\S\ref{sec:keynotes}), followed by summaries of the
workshop papers grouped by the three thematic workshop panels to which
they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Two
broader cross-cutting issues surfaced repeatedly, and are discussed
separately~(\S\ref{sec:cross-cutting}), as are use cases for
sustainable software~(\S\ref{sec:use-cases}). The summaries are based
not only on the papers and panel presentations, but also on the many
comments raised in both the onsite and online discussions that
accompanied the workshop elements, as documented by collaborative note
taking during the workshop~\cite{WSSSPE1-google-notes}. We conclude
with issues and lessons learned, as well as plans for future
activities~(\S\ref{sec:conclusions}).  The short papers accepted to
the workshop are listed in Appendix~\ref{sec:papers}, and a partial
list of workshop attendees can be found in
Appendix~\ref{sec:attendees}. The original call for papers is included in 
Appendix~\ref{sec:cfp}.

@@

\subsubsection{Licensing and Software Patents}
Software is no longer either open or closed (only binaries available),
but also licensed and patented, which clearly also impact 
software sustainability.
While many papers briefly discussed licensing issues,
Elster~\cite{Elster_WSSSPE} directly discussed the impact of software licenses 
on obtaining industrial funding for scientific software
projects.  In particular, she describes her experiences with researchers
unwilling to utilize GPL (copyleft) code, as it adds restrictions to
reuse that they themselve as well as some industries find unacceptable for future
commercialization.  (This was discussed by Hanwell et
al.~\cite{Hanwell_WSSSPE} as well.)  US ICT companies funding 
academics will thus often insist on BSD licensing on software 
since they then can legally include the code into their commercial
codes.

On the other hand, there are companies
that fund larger GPL-licensed software projects~\cite{Blatt_WSSSPE}
and insist that the university projects they fund 
also produce code with GPL licensing.
They do not accept BSD-like licenses since they argue that
other companies then may choose to build closed commercial
codes on they software they funded, rather than encouraging 
the community to contribute freely and thus ensuring software 
sustainability for the community.

In either case, the university researchers are not given much choice
if they want these much sought after funds in a increasingly competitive
grant world.

Another obstacles to
sustainability identified by the author include patenting of software.
Most countries place some limits on software patents. The European 
Union outright forbids them, while US patent law excludes “abstract ideas” 
that has been used to refuse some software patents.

Further obstacles to
sustainability is a lack of open access, and even more broadly, a lack of open source
codes even in open access journals.  Finally, a lack of awareness on
the part of scientific software developers of commodity libraries for
common tasks reduces their ability to reuse code.

@@

Two issues, how to define software sustainability, and how career
paths (or their lack) interact with achieving it, were raised across
the workshop's panels.  This section aims to synthesize these
discussions from different perspectives.

@@

\note{add a subsection about licensing - lessons, advice, etc.?}

@@

Three issues came up repeatedly in different parts of the workshop:

@@

@@

\end{enumerate}
In addition, licensing and patents and how they impact research funding
for software developement, was also discussed.
% Conclusion of S8 : Use Cases, not sure where to put
%\note{conclusions from S8, not sure of location --Ketan}

>>>>

This report attempts to summarize the various aspects of the workshop.
The remainder of this paper first gives an overview of the process
with which the workshop was put together (\S\ref{sec:process}), then
proceeds with summaries of the two keynote
presentations~(\S\ref{sec:keynotes}), followed by summaries of the
workshop papers grouped by the three thematic workshop panels to which
they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Three
broader cross-cutting issues surfaced repeatedly, and are discussed
separately~(\S\ref{sec:cross-cutting}), as are use cases for
sustainable software~(\S\ref{sec:use-cases}). The summaries are based
not only on the papers and panel presentations, but also on the many
comments raised in both the onsite and online discussions that
accompanied the workshop elements, as documented by collaborative note
taking during the workshop~\cite{WSSSPE1-google-notes}. We conclude
with issues and lessons learned, as well as plans for future
activities~(\S\ref{sec:conclusions}).  The short papers accepted to
the workshop are listed in Appendix~\ref{sec:papers}, and a partial
list of workshop attendees can be found in
Appendix~\ref{sec:attendees}. The original call for papers is included in 
Appendix~\ref{sec:cfp}.

@@

@@

Three issues, how to define software sustainability, how career
paths (or their lack) interact with achieving it, and the impact of
software licenses, were raised across
the workshop's panels.  This section aims to synthesize these
discussions from different perspectives.

@@

\subsection{Licensing and Software Patents}
Issues related to licensing and patents primarily was discussed in the Communities
session, but licensing was also a concern of many of the other contributors
in other sessions.
%
Software is no longer just open or closed (only binaries available),
but also licensed and patented, which clearly also impacts 
software sustainability.
While many papers briefly discussed licensing issues,
Elster~\cite{Elster_WSSSPE} directly discussed the impact of software licenses 
on obtaining industrial funding for scientific software
projects.  In particular, she described her experiences with researchers
unwilling to utilize GPL (copyleft) code, as it adds restrictions to
reuse that they themselves as well as some industries find unacceptable for future
commercialization.  (This was discussed by Hanwell et
al.~\cite{Hanwell_WSSSPE} as well.)

US information technology companies funding 
academics will thus often insist on BSD licensing on software 
since they then can legally include the code into their commercial
codes.
%
On the other hand, there are companies
that fund larger GPL-licensed software projects~\cite{Blatt_WSSSPE}
and insist that the university projects they fund 
also produce code with GPL licensing.
They do not accept BSD-like licenses since they argue that
other companies then may choose to build closed commercial
codes on they software they funded, rather than encouraging 
the community to contribute freely and thus ensuring software 
sustainability for the community.
%
In either case, the university researchers are not given much choice
if they want these much sought after funds in a increasingly competitive
grant world.

Another obstacle to
sustainability identified by Elster include patenting of software.
Most countries place some limits on software patents. The European 
Union outright forbids them, while US patent law excludes ``abstract ideas,'' 
which has been used to refuse some software patents.
%
Further obstacles to
sustainability include a lack of open access, and even more broadly, a lack of open source
codes even in open access journals.  Finally, a lack of awareness on
the part of scientific software developers of commodity libraries for
common tasks reduces their ability to reuse code.

@@

Three major issues came up repeatedly in different parts of the workshop:

@@

\end{enumerate}
%
In addition, licensing and patents, and how they impact research funding
for software development, were also discussed.

@@

% Conclusion of S8 : Use Cases, not sure where to put
%\note{conclusions from S8, not sure of location --Ketan}
